---
title: BIOS 7323 Lab 11
date: 2018-11-26
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=file.path("/home/nathan/Dropbox/njames/school/PhD/courses/2018_19/ta_surv_anlys2/labs/lab11"))

#tidyverse pkgs
library(tidyverse)
library(magrittr)

#survival analysis pkgs
library(survival)
library(SurvRegCensCov)
```

### Objectives

Understand different representations of (Weibull) parametric survival analyses

### Data

The file liver.txt contains the data for 418 patients with primary biliary cirrhosis, with stage information available for 312 patients. Dataset obtained from emersonstatistics.com.

### Parametric model forms

Let $X$ be the time to event and $Z$ a vector of fixed time explanatory variables.

> *Accelerated failure time (AFT) form*

$S(x|Z)=S_0(\exp(\theta'Z)x)$ 

where $\theta'=(\theta_1,\ldots,\theta_p)$ is a vector of regression coefficients

Survival of an individual with covariates $Z$ at time $x$ is the same as baseline survival at time $\exp(\theta'Z)x$. The term $\exp(\theta'Z)$ is called the acceleration factor.

> *Log-linear form*

In the log-linear form, if $W \sim$ standard extreme value distribution, then $X \sim$ Weibull. The model in this form is:

$\log X = Y = \mu +\gamma'z +\sigma W$ 

where $\gamma'=(\gamma_1,\ldots,\gamma_p)$ is a vector of regression coefficients and $W$ is the error random variable

If $S_0(x)$ is the survival function of the random variable $\exp(\mu+\sigma W)$ then the AFT and log-linear forms are equivalent with $\theta=-\gamma$. This means we can estimate parameters in the log-linear form and transform to interpret the covariates in AFT form. 

> *Proportional hazards form*

 The Weibull model also has a proportion hazards (PH) form:

$h(x|Z)=h_0(x)\exp(\beta'Z)=\alpha \lambda x^{\alpha-1}\exp(\beta'Z)$

The PH form is related to the log-linear form through the reparametrization:

$\alpha =\frac{1}{\sigma} \qquad \lambda=\exp(\frac{-\mu}{\sigma}) \qquad \beta_j=\frac{-\gamma_j}{\sigma},\; j=1,\ldots,p$

```{r}
#load the data 
liver_dat<-read_table("liver.txt")

#keep only stage, death status, and observed time variables and remove those missing stage
liver<-select(liver_dat, obstime, status, stage) %>% remove_missing()

# fit a parametric Weibull regression with stage as the covariate using survreg(..., dist="weibull")
mod1 <- survreg(Surv(obstime,status) ~ factor(stage), data=liver, dist="weibull")

# Find the MLE of median survival for a patient with stage 1 primary biliary cirrhosis using predict(..., type="quantile", p=0.5, newdata=data.frame(stage=1)) 


# What is the MLE of median survival for a patient with stage 4 primary biliary cirrhosis? 


# get a summary of the model fit, note that the parameters are in the log-linear form
summary(mod1)

# using the output from survreg() extract and save the estimates of mu (Intercept) and sigma (Scale)


# Now get the estimates for var(mu), var(sigma) and cov(mu,sigma)
# Note: the variance/covar matrix from survreg() has the variance of log(sigma), not sigma
# using delta-method:
#   var(sigma)=var(log(sigma))*sigma^2
#   cov(mu,sigma)=cov(mu,log(sigma))*sigma


```

```{r}
## Proportional Hazards form ##

# using the output from survreg() and the reparameterization above, get point estimates for
# lambda and alpha in the PH form


# now get the variance of lambda and alpha using K&M formulas (12.2.14) and (12.2.15)

#12.2.14
# varlambda=exp(-2 * (mu/sigma)) * ((varmu/sigma^2) - (2 * (mu * cvmusigma)/sigma^3) + ((mu^2 * varsigma)/sigma^4)) 

#12.2.15
# varalpha=varsigma/sigma^4 
```


```{r}
## Accelerated failure time form ##

# using the output from survreg() and the appropriate reparameterization, get the acceleration factors for stage 2, 3, and 4 compared to stage 1. 

# How do you interpret the acceleration factor?

```

```{r}
## Use WeibullReg() in the SurvRegCensCov package to get parameter estimates in PH form and AFT form
# and verify they match those calculated above

# Note: WeibullReg() uses the same parameterizations as K&M, but switches the names of some variables. For example, alpha=1/sigma in K&M is called gamma by WeibullReg(). see vignette("weibull")

# Hint: In WeibullReg() output 1/ETR = Acceleration factor

```

#### Parametric Model Diagnostics

Cox-Snell residuals can be used to check the overall fit of the model. For the Weibull model the C-S residuals are:

$r_i = \hat{\lambda}\exp(\hat{\beta}'\mathbf{Z}_i)t_{i}^{\hat{\alpha}}$

If the model fits the data, a plot of $r_i$ vs. the Nelson-Aalen estimator of the cumulative hazard of the $r_i$'s should be a straight line with slope 1.

```{r}
# assess the overall fit of the Weibull model by producing a cumulative hazard plot for the Cox-Snell residuals (see K&M p. 414)
# Hint: first reparameterize from the log-linear model gamma parameters to PH beta parameters

# Does the Weibull model fit well based on the C-S cumulative hazard plot?


```
