---
title: BIOS 7323 Lab 9
date: 2018-11-05
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=file.path("/home/nathan/Dropbox/njames/school/PhD/courses/2018_19/ta_surv_anlys2/labs/lab9"))

#tidyverse pkgs
library(tidyverse)
library(magrittr)

#survival analysis pkgs
library(survival)
```

### Objectives {.tabset}

Understand the proportional hazards assumption in Cox regression

#### Data

The file liver.txt contains the data for 418 patients with primary biliary cirrhosis. 312 of these patients were randomized to receive either D-penicillamine or placebo in a clinical trial. Limited data is available on the other 106 patients. Each row of the file corresponds to a single patient. (The first row of the file contains the variable names.) Dataset obtained from emersonstatistics.com.

For this lab, we will investigate whether stage of liver disease is associated with survival. Note: For simplicity, we will not adjust for other variables, but this is not recommended in practice.

#### Visualize proportional hazards from Cox model 

```{r}
#load the data 
liver_dat<-read_table("liver.txt")

#keep only stage, death status, and observed time variables
liver<-select(liver_dat, obstime, status, stage)

#Fit a Cox PH model for the association between time to death and disease stage
# including disease stage as a CATEGORICAL variable
modelA <- coxph(Surv(obstime,status)~factor(stage), data=liver)

# Plot the predicted survival curves for each stage based on the cox PH model
plot(survfit(modelA, newdata=data.frame(stage=1:4)), col=1:4)

#Use basehaz() to produce estimates of the cumulative hazard
# when instant_haz() is used below, basehaz() is called to get these cumulative hazards
cumhaz<-survival::basehaz(modelA, center=FALSE)

#load the supplemental function
source("inst_haz.R")

#use instant_haz(coxph.fit) to produce estimates of the baseline instantaneous hazard
instanthaz_A <- instant_haz(modelA)

# plot the instantaneous hazard over time for each stage
# hint: multiply the baseline hazard by the appropriate hazard ratio from the model to get estimates for stage 2, 3, and 4

all_inst_hazA<-ggplot(data=instanthaz_A, aes(x=time,y=inst_haz))+
  geom_line()+
  geom_line(aes(x=time,y=inst_haz*exp(coef(modelA)[1])), color=2)+
  geom_line(aes(x=time,y=inst_haz*exp(coef(modelA)[2])), color=3)+
  geom_line(aes(x=time,y=inst_haz*exp(coef(modelA)[3])), color=4)+
  ylab("inst. hazard")

all_inst_hazA

# plot the log instantaneous hazard over time for each stage
# Note: this code is actually transforming the y-axis NOT the actual hazard values
all_inst_hazA+scale_y_continuous(trans="log")+ylab("log inst. hazard")

# plot a smoothed estimate of the log instantaneous hazards for each stage
all_inst_hazA_2<-ggplot(data=instanthaz_A, aes(x=time,y=inst_haz))+
  geom_smooth(method="loess", se=FALSE, color=1)+
  geom_smooth(aes(x=time,y=inst_haz*exp(coef(modelA)[1])), color=2, method="loess", se=FALSE)+
  geom_smooth(aes(x=time,y=inst_haz*exp(coef(modelA)[2])), color=3, method="loess", se=FALSE)+
  geom_smooth(aes(x=time,y=inst_haz*exp(coef(modelA)[3])), color=4, method="loess", se=FALSE)+
  scale_y_continuous(trans="log")+ylab("smoothed log inst. hazard")

all_inst_hazA_2

if (0){
#smooth log inst. hazard using base R
inst_haz_loess<-loess.smooth(instanthaz_A$time,instanthaz_A$inst_haz)

plot(inst_haz_loess$x, log(inst_haz_loess$y),type="l",ylim=c(-5,-1))
lines(inst_haz_loess$x, log(inst_haz_loess$y*exp(coef(modelA)[1])),type="l", col=2)
lines(inst_haz_loess$x, log(inst_haz_loess$y*exp(coef(modelA)[2])),type="l", col=3)
lines(inst_haz_loess$x, log(inst_haz_loess$y*exp(coef(modelA)[3])),type="l", col=4)
}

# Do the hazards appear to be proportional?

```

```{r}

#Fit a Cox PH model for the association between disease stage and time to death 
# including disease stage as a CONTINUOUS variable
modelB <- coxph(Surv(obstime,status)~stage, data=liver)

# Plot the predicted survival curves for each stage based on the cox PH model
plot(survfit(modelB, newdata=data.frame(stage=1:4)), col=1:4)

#Use basehaz() to produce estimates of the cumulative hazard
# when instant_haz() is used below, basehaz is called to get these cumulative hazards
cumhazB<-survival::basehaz(modelB, center=FALSE)

#use instant_haz(coxph.fit) to produce estimates of the baseline instantaneous hazard
instanthaz_B <- instant_haz(modelB)

# plot a smoothed estimate of the log instantaneous hazards for each stage
all_inst_hazB_2<-ggplot(data=instanthaz_B, aes(x=time,y=inst_haz*exp(1*coef(modelB))))+
  geom_smooth(method="loess", se=FALSE, color=1)+
  geom_smooth(aes(x=time,y=inst_haz*exp(2*coef(modelB))), color=2, method="loess", se=FALSE)+
  geom_smooth(aes(x=time,y=inst_haz*exp(3*coef(modelB))), color=3, method="loess", se=FALSE)+
  geom_smooth(aes(x=time,y=inst_haz*exp(4*coef(modelB))), color=4, method="loess", se=FALSE)+
  scale_y_continuous(trans="log")+ylab("smoothed log inst. hazard")

all_inst_hazB_2

# What is the difference between the first and second models in terms of the PH assumption?

```



```{r}

#Fit a Cox PH model for time to death stratified by stage 
modelC<-coxph(Surv(obstime,status) ~ strata(stage), data=liver)

# produce the instantaneous hazards for each strata using instant_haz(coxph.fit)
instanthaz_C<-instant_haz(modelC)

# plot the smoothed log instantaneous hazards for each strata
base_inst_hazC <- ggplot(data=instanthaz_C, aes(x=time, y=inst_haz, color=strata))+
                  geom_smooth(method="loess", se=FALSE)+
                  scale_y_continuous(trans="log")+
                  ylab("smoothed log inst. hazard")

base_inst_hazC

# Does the proportional hazards assumption appear to hold between strata

```
