---
title: BIOS 7323 Lab 9
date: 2018-11-05
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=file.path("/home/nathan/Dropbox/njames/school/PhD/courses/2018_19/ta_surv_anlys2/labs/lab9"))

#tidyverse pkgs
library(tidyverse)
library(magrittr)

#survival analysis pkgs
library(survival)
```

### Objectives

Understand the proportional hazards assumption in Cox regression

#### Data

The file liver.txt contains the data for 418 patients with primary biliary cirrhosis. 312 of these patients were randomized to receive either D-penicillamine or placebo in the clinical trial. Limited data is available on the other 106 patients. Each row of the file corresponds to a single patient. (The first row of the file contains the variable names.) Dataset obtained from emersonstatistics.com.

For this lab, we will investigate whether stage of liver disease is associated with survival. Note: For simplicity, we will not adjust for other variables, but this is not recommended in practice.

#### Visualize proportional hazards from Cox model

```{r}
# load the data 
liver_dat<-read_table("liver.txt")

# keep only stage, death status, and observed time variables
liver<-select(liver_dat, obstime, status, stage)

#Fit a Cox PH model for the association between time to death and disease stage
# including disease stage as a CATEGORICAL variable


# Plot the predicted survival curves for each stage based on the cox PH model


# load the supplemental function
source("inst_haz.R")

# use instant_haz(coxph.fit) to produce estimates of the baseline instantaneous hazard


# plot the instantaneous hazard over time for each stage
# hint: multiply the baseline hazard by the appropriate hazard ratio from the model to get estimates for stage 2, 3, and 4


# plot a smoothed estimate of the log instantaneous hazards for each stage


# Do the hazards appear to be proportional?


```



```{r}
#Now fit a Cox PH model for the association between disease stage and time to death 
# including disease stage as a CONTINUOUS variable


# Plot the predicted survival curves for each stage based on the cox PH model


# use instant_haz(coxph.fit) to produce estimates of the baseline instantaneous hazard


# plot a smoothed estimate of the log instantaneous hazards for each stage


# What is the difference between the first and second models in terms of the PH assumption?
```


```{r}
#Fit a Cox PH model for time to death stratified by stage 


# produce the instantaneous hazards for each strata using instant_haz(coxph.fit)


# plot the smoothed log instantaneous hazards for each strata


# Does the proportional hazards assumption appear to hold between strata?

```

#### Graphical checks of Proportional Hazards Assumption

There are several approaches to graphically check the proportional hazards assumption. Here we will examine an approach based on cumulative hazard estimates from a stratified Cox model (See K&M Chp 11.4 and pgs. 19-22 in Chp 11 lecture notes). If the proportional hazards assumption holds, the baseline cumulative hazard rates in each strata should be a constant multiple of each other.

```{r}
#using the model stratified by stage calculate the cumulative hazards using basehaz()
# and plot the log cumulative hazards vs. time


```

```{r}
#plot the differences in log cumulative hazard (e.g. log(stage1)-log(stage4), log(stage2)-log(stage4), etc.) vs. time


```
 
```{r}
#produce the Andersen plot (cumhaz for stage 1 vs. cumhaz for stage 4, cumhaz for stage 2 vs. cumhaz for stage 4, etc.)


# Based on these three plots, does disease stage meet the PH assumption?

```
