---
title: BIOS 7323 Lab 8
date: 2018-10-29
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=file.path("/home/nathan/Dropbox/njames/school/PhD/courses/2018_19/ta_surv_anlys2/labs/lab8"))

#tidyverse pkgs
library(tidyverse)
library(survminer)
library(magrittr)
library(broom)

#survival analysis pkgs
library(survival)
library(KMsurv)
```

### Objectives

Code time-dependent covariates

Check PH assumptions using time-dependent covariates

Use Stratified PH models

#### Data

The file heart_trans.csv contains data for 102 patients on a waiting list for heart transplant.

id: subject id  
delta: death indicator (0=alive, 1=died)  
futime: time from acceptance on wait list to death or end of follow-up  
surgery: prior bypass surgery indicator  
age: centered age in years at acceptance on wait list (i.e., age-48)  
trx.wait.time: time before transplant  
trx: heart transplant indicator  
era: indicator for transplant era (before or after 1971)  

The primary research question is whether prior bypass surgery affects overall survival after acceptance onto the wait list. 

#### Coding time-dependent covariates

Because some patients undergo heart transplantation, it is also of interest to see how this time-dependent covariate affects overall survival. We will explore the effect of transplantation using a time-transform function `tt()` and by creating a counting process dataset.

```{r}
#load the data from heart_trans.csv
heart_dat<-read_csv("heart_trans.csv",col_types = c("iiilddli"))

#create a variable which is equal to the time before transplant (trx.wait.time) for those who received a transplant and is greater than the maximum follow-up time otherwise
heart_dat %<>% mutate(ttrans=if_else(trx,trx.wait.time,max(futime)+1))

#using the variable you just created and tt() fit a Cox PH model that also includes surgery and age
ttfit<-coxph(Surv(futime,delta) ~ surgery + age + tt(ttrans), data=heart_dat,
             tt=function(x,t,...) (t>=x))

# what is the impact of prior bypass surgery on overall survival?


# now fit the same model by creating a counting process dataset 

# using tmerge()

heart_cp_dat <- tmerge(data1=heart_dat, data2=heart_dat, id=id, death=event(futime,delta)) %>%
                tmerge(data1=.,data2=.,id=id,Ztrans=tdc(ttrans-0.5))

heart_cp_dat

tmergefit <- coxph(Surv(tstart,tstop,death) ~ surgery + age + Ztrans, data=heart_cp_dat)

#  using survSplit()

#unique transplant times (w/ end shifted)
ttimes <- heart_dat %>% select(ttrans) %>% unique() %>% pull() - 0.5

heart_cp_dat2 <- survSplit(Surv(futime,delta) ~ trx.wait.time + surgery + age, cut=ttimes,
                           id="id", data=heart_dat) %>% 
                 mutate(ttrx=if_else(is.na(trx.wait.time),FALSE,futime>=trx.wait.time))

heart_cp_dat2
survsplitfit <- coxph(Surv(tstart,futime,delta) ~ surgery + age + ttrx, data=heart_cp_dat2)

# how many rows are in your counting process dataset?
nrow(heart_cp_dat)
nrow(heart_cp_dat2)

heart_dat %>% filter(id %in% 1:4)
heart_cp_dat %>% filter(id %in% 1:4)
heart_cp_dat2 %>% filter(id %in% 1:4)

# how do you interpret the coefficient for transplantation?

( td_mods <- map(list(ttfit,tmergefit,survsplitfit),function(x) tidy(x,exponentiate=TRUE)) )

# --> At a given time, a person with transplant has better survival than person w/o transplant
# --> estimate for hazard of death after transplant is 0.964 hazard of death before transplant, but not statistically significant

```


#### Checking proportional hazards assumptions using time-dependent covariates

A time dependent covariate can be used to test the PH assumption for a fixed-time covariate $Z_1$ by using a known function `g(t)` to create a artifical time-dependent variable $Z_2(t)=Z_1 \times g(t)$. The model for the time-dependent covariate is:

$\lambda(t)=\lambda_0(t)\exp(\beta Z(t))$

A test of $H_0:\beta_2=0$ is a test of the PH assumption for the fixed-time covariate.

Alternatively, we can consider a time-dependent *coefficient* in the model:

$\lambda(t)=\lambda_0(t)\exp(\beta(t) Z)$

Note that in this model, the $\beta$ are a function of time and the $Z$ are constant. If the PH assumption is true $\beta(t)$ will be a horizontal line (i.e. slope = 0).

See the `Using Time Dependent Covariates` vignette in the `survival` package for more info.

```{r}

ggsurvplot( survfit(Surv(futime,delta)~surgery,data=heart_dat))

#test whether the PH assumption for surgery holds by fitting a Cox PH model with age, surgery, and the artifical time-dependent variable surgery x log(time) using tt()

fit1<-coxph(Surv(futime,delta) ~ age + surgery + tt(surgery),
            tt=function(x,t,...) x*log(t), data=heart_dat)
summary(fit1)

# Is the PH assumption is violated for surgery?


#fit a cox PH model with fixed covariates age and surgery
fit0 <- coxph(Surv(futime,delta)~ age + surgery, data=heart_dat)

#use this model fit in cox.zph() to estimate time dependent coefficients, beta(t), using a log transformation

( phtest <- cox.zph(fit0,transform='log') )

# equivalent
# cox.zph(fit0,transform=function(futime) log(futime))

#plot the time-dependent coefficient for surgery
plot(phtest[2])
abline(0,0,col=2) # line for no effect
abline(h=fit0$coef[2],col=3,lty=2) # line for avg hazard over time from PH model

# based on the test and plot of the time-dependent coefficient, does the PH assumption hold for surgery?
```

#### Stratified proportional hazards model

One approach to correct for a covariate which violated the proportional hazards assumption is to stratify on the covariate and fit a PH model for the other covariates within each strata, $j=1,\ldots,s$.

$\lambda_j(t)=\lambda_{0j}(t)\exp(\beta Z(t))$

The baseline hazards are allowed to vary across strata, but the coefficients are assumed to be the same for each strata. We can test this assumption using a likelihood ratio test or a Wald test.

```{r}
# fit a Cox PH model with fixed surgery and age covars, a time-dependent covariate for transplant, and a strata for each era
erafit_str<-coxph(Surv(futime,delta)~surgery+age+tt(ttrans)+strata(era),data=heart_dat,
                  tt=function(x,t,...) (t>=x))

#fit the model above separately for the patients in each era
erafit_0<-coxph(Surv(futime,delta)~surgery+age+tt(ttrans),data=subset(heart_dat,era==0),
                tt=function(x,t,...) (t>=x))

erafit_1<-coxph(Surv(futime,delta)~surgery+age+tt(ttrans),data=subset(heart_dat,era==1),
                tt=function(x,t,...) (t>=x))

# perform a likelihood ratio test for the null hypothesis of equal effects across strata 
lrt <- -2*(logLik(erafit_str) - (logLik(erafit_0) + logLik(erafit_1)))

# compare to chi-sq with (s-1)*p=(2-1)*3 df
pchisq(lrt,df=(2-1)*3,lower.tail = F)


# perform a Wald test of the same hypothesis

#coefficients from models in 2 strata
(strcoef<-c(erafit_0$coef, erafit_1$coef))

#block diagonal covar matrix
offdiag<-matrix(rep(0,9),nrow=3)
( strvar<-rbind(cbind(erafit_0$var,offdiag),cbind(offdiag,erafit_1$var)) )

#contrast matrix
C=matrix(c(1,0,0,-1,0,0,
         0,1,0,0,-1,0,
         0,0,1,0,0,-1),
         nrow=3,byrow=TRUE)

# wald test statistic
wald<-t(C%*%strcoef) %*% solve(C %*% strvar %*% t(C)) %*% (C%*%strcoef)

pchisq(wald,3,lower.tail=FALSE)

# What conclusion can you draw from these tests?

# --> not enough evidence to detect difference in covariate effects for different eras
```


```{r,eval=FALSE,echo=FALSE}
#code to make heart_trans

# data(heart)

#keep only need vars, drop pt who died on same day as transplant to avoid (0,0) interval
heart_dat<-jasa %>% as.tibble() %>% mutate(id=1:n()) %>% 
  mutate(age=age-48,era=if_else(accept.dt>=as.Date("1971-01-01"),1,0)) %>%
  select(id,fustat,futime,surgery,age,wait.time,transplant,era) %>%
  rename(delta=fustat,trx.wait.time=wait.time,trx=transplant) %>%
  filter(id!=15) 

write_csv(heart_dat,file.path(getwd(),"heart_trans.csv"))
```
