---
title: BIOS 7323 Lab 8
date: 2018-10-29
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=file.path("/home/nathan/Dropbox/njames/school/PhD/courses/2018_19/ta_surv_anlys2/labs/lab8"))

#tidyverse pkgs
library(tidyverse)
library(survminer)
library(magrittr)
library(broom)

#survival analysis pkgs
library(survival)
library(KMsurv)
```

### Objectives

Code time-dependent covariates

Check PH assumptions using time-dependent covariates

Use Stratified PH models

#### Data

The file heart_trans.csv contains data for 102 patients on a waiting list for heart transplant.

id: subject id  
delta: death indicator (0=alive, 1=died)  
futime: time from acceptance on wait list to death or end of follow-up  
surgery: prior bypass surgery indicator  
age: centered age in years at acceptance on wait list (i.e., age-48)  
trx.wait.time: time before transplant  
trx: heart transplant indicator  
era: indicator for transplant era (before or after 1971)  

The primary research question is whether prior bypass surgery affects overall survival after acceptance onto the wait list. 

#### Coding time-dependent covariates

Because some patients undergo heart transplantation, it is also of interest to see how this time-dependent covariate affects overall survival. We will explore the effect of transplantation using a time-transform function `tt()` and by creating a counting process dataset.

```{r}
#load the data from heart_trans.csv
heart_dat<-read_csv("heart_trans.csv",col_types = c("iiilddli"))

# create a variable which is equal to the time before transplant (trx.wait.time) for those who received a transplant and is greater than the maximum follow-up time otherwise
heart_dat %<>% mutate(ttrans=if_else(trx,trx.wait.time,max(futime)+1))

#using the variable you just created and tt() fit a Cox PH model that also includes surgery and age
ttfit<-coxph(Surv(futime,delta)~surgery+age+tt(ttrans),data=heart_dat,
             tt=function(x,t,...) (t>=x))

# what is the impact of prior bypass surgery on overall survival in this model?


# now fit the same model by creating a counting process dataset 
# e.g. use tmerge() or survSplit()



# how many rows are in your counting process dataset?


# how do you interpret the coefficient for transplantation?


```


#### Checking proportional hazards assumptions using time-dependent covariates

A time dependent covariate can be used to test the PH assumption for a fixed-time covariate $Z_1$ by using a known function `g(t)` to create a artifical time-dependent variable $Z_2(t)=Z_1 \times g(t)$. The model for the time-dependent covariate is:

$\lambda(t)=\lambda_0(t)\exp(\beta Z(t))$

A test of $H_0:\beta_2=0$ is a test of the PH assumption for the fixed-time covariate.

Alternatively, we can consider a time-dependent *coefficient* in the model:

$\lambda(t)=\lambda_0(t)\exp(\beta(t) Z)$

Note that in this model, the $\beta$ are a function of time and the $Z$ are constant. If the PH assumption is true $\beta(t)$ will be a horizontal line (i.e. slope = 0).

See the `Using Time Dependent Covariates` vignette in the `survival` package for more info.

```{r}
#test whether the PH assumption for surgery holds by fitting a Cox PH model with age, surgery, and the artifical time-dependent variable surgery x log(time) using tt()


#Is the PH assumption is violated for surgery?


#fit a cox PH model with fixed covariates age and surgery


#use this model fit in cox.zph() to estimate time dependent coefficients, beta(t), using a log transformation


#plot the time-dependent coefficient for surgery using the output of cox.zph


# based on the test and plot of the time-dependent coefficient, does the PH assumption hold for surgery?
```

#### Stratified proportional hazards model

One approach to correct for a covariate which violated the proportional hazards assumption is to stratify on the covariate and fit a PH model for the other covariates within each strata, $j=1,\ldots,s$.

$\lambda_j(t)=\lambda_{0j}(t)\exp(\beta Z(t))$

The baseline hazards are allowed to vary across strata, but the coefficients are assumed to be the same for each strata. We can test this assumption using a likelihood ratio test or a Wald test.

```{r}
# fit a Cox PH model with fixed surgery and age covars, a time-dependent covar for transplant, and a strata for each era


#fit the model above separately for the patients in each era



# perform a likelihood ratio test for the null hypothesis of equal effects across strata 
# lrt <- -2*(logLik(fit_stratified) - (logLik(fit_strata_0) + logLik(fit_strata_1)))

# compare to chi-sq with (s-1)*p=(2-1)*3 df
# pchisq(lrt, df=(2-1)*3, lower.tail = FALSE)

# perform a Wald test of the same hypothesis

#coefficients from models in 2 strata
# strcoef<-c(fit_strata_0$coef, fit_strata_1$coef) 

#make block diagonal covar matrix
# offdiag<-matrix(rep(0,9),nrow=3)
# strvar<-rbind(cbind(fit_strata_0$var,offdiag),cbind(offdiag,fit_strata_1$var)) )

#contrast matrix
# C=matrix(c(1,0,0,-1,0,0,
#         0,1,0,0,-1,0,
#         0,0,1,0,0,-1),
#         nrow=3,byrow=TRUE)

# get wald test statistic


# What conclusion can you draw from these tests?

```
