---
title: BIOS 7323 Lab 4
date: 2018-09-24
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=file.path("/home/nathan/Dropbox/njames/school/PhD/courses/2018_19/ta_surv_anlys2/labs/lab4"))

#tidyverse pkgs
library(tidyverse)
library(ggfortify)
library(magrittr)

#survival analysis pkgs
library(survival)
library(KMsurv)
```

### Objectives

Calculate point and interval estimates of mean, median, and other quantiles of survival time

Estimate survival for left truncated, right censored data

#### Point estimates of S(t) and survival times

```{r}
# Load the tongue cancer data set and use it to answer the following questions
data(tongue)

tongue_km <- survfit(Surv(time, delta) ~ 1, type="kaplan-meier", conf.type="log-log", data=tongue)

# !!! NOTE !!! 
# std.err from survfit.object is the standard error of the cumulative hazard function
# std.err from summary.survfit is the standard error of survival function
# see ?survfit.object and ?summary.survfit

# What is the median survival time?


# What is the probability of survival at 26 weeks? 52 weeks?


# At what time do we estimate 75% of patients will have died? 90%?


# At what time are 10% (or less) of the patients still in the study?


```

#### Estimates of mean and confidence interval

In K & M Chapter 2, it was shown that the mean time to event is

$\mu = \int_{0}^{\infty} S(t)dt$

An estimate of the mean survival can be obtained by plugging in the nonparametric estimate of $S(t)$

$\hat{\mu} = \int_{0}^{\tau} \hat{S(t)}dt$

where $\tau$ must be selected by the investigator.

```{r}
# Use the survfit object and the print method with option rmean="common" to estimate the mean survival time and a 95% confidence interval using Efron's tail correction


# Estimate the mean survival time and a 95% confidence interval assuming no one can survive past 300 weeks


# Which estimate do you prefer? Why?

```

#### Confidence interval for quantiles

```{r}
# Calculate a 95% log-log CI for the median survival time using the method in K & M Chp. 4 (p. 120-122)


# compare this to the 95% CI for the median that is produced when printing a summary of the survfit object (or using quantile.survfit). Are the intervals the same?


```


#### Estimators for left-truncated, right censored data

```{r}
# read in the psych data and create a survival object using 'ageentry' as the left truncation time and 'agestop' as the right censoring time
psych <- read_csv("psych.csv")

# get K-M estimates and log-log confidence intervals by sex given survival to age 21

            
#plot conditional survival estimates given survival to age 21


#plot conditional survival estimates given survival to age 35


```
