---
title: BIOS 7323 Lab 4
date: 2018-09-24
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=file.path("/home/nathan/Dropbox/njames/school/PhD/courses/2018_19/ta_surv_anlys2/labs/lab4"))

#tidyverse pkgs
library(tidyverse)
library(ggfortify)
library(magrittr)

#survival analysis pkgs
library(survival)
library(KMsurv)
```

### Objectives

Calculate point and interval estimates of mean, median, and other quantiles of survival time

Estimate survival for left truncated, right censored data

#### Point estimates of S(t) and survival times

```{r}
# Load the tongue cancer data set and use it to answer the following questions
data(tongue)

tongue_km <- survfit(Surv(time, delta)~1, type="kaplan-meier",conf.type="log-log", data=tongue)

surv_tab <- summary(tongue_km) %$% cbind(time, n.risk, n.event, n.censor, surv, std.err) %>% as.tibble()

# !!! NOTE !!! 
# std.err from survfit.object is the standard error of the cumulative hazard function
# std.err from summary.survfit is the standard error of survival function
# see ?survfit.object and ?summary.survfit

# What is the median survival time?

print(tongue_km)
#  73 weeks

# What is the probability of survival at 26 weeks? 52 weeks?

surv_tab %>% filter(time>=26)
surv_tab %>% filter(time>=51)

#  72.3% and 59.6%,respectively


# At what time do we estimate 75% of patients will have died? 90%?

surv_tab %>% filter(surv<0.25)
#  inf(t:S(t)<=1-p)
#  167 weeks is time when >75% of patients will have died
#  survival curve never goes below 10%, so no estimate for time when 90% will have died

# At what time are 10% (or less) of the patients still in the study?

surv_tab %>% filter(n.risk<=8)
#  look at n.risk 8 pts at risk at time 157

```

#### Estimates of mean and confidence interval

In K & M Chapter 2, it was shown that the mean time to event is

$\mu = \int_{0}^{\infty} S(t)dt$

An estimate of the mean survival can be obtained by plugging in the nonparametric estimate of $S(t)$

$\hat{\mu} = \int_{0}^{\tau} \hat{S(t)}dt$

where $\tau$ must be selected by the investigator.

```{r}
# Use the survfit object and the print method with option rmean="common" to estimate the mean survival time and a 95% confidence interval using Efron's tail correction

print(tongue_km, rmean="common")
print(tongue_km, rmean=max(tongue$time))

tab_efron <- tongue_km %>% summary(rmean="common") %>% extract2("table")
mn_ef <- tab_efron %>% extract("*rmean")
se_ef <- tab_efron %>% extract("*se(rmean)")

mn_ef + c(-1,1)*qnorm(0.975)*se_ef

# Estimate the mean survival time and a 95% confidence interval assuming no one can survive past 300 weeks

print(tongue_km, rmean=300)

tab_300<-tongue_km %>% summary(rmean=300) %>% extract2("table")
mn_300<-tab_300 %>% extract("*rmean")
se_300<-tab_300 %>% extract("*se(rmean)")

mn_300 + c(-1,1)*qnorm(0.975)*se_300

# Which estimate do you prefer? Why?

```

#### Confidence interval for quantiles

```{r}
# Calculate a 95% log-log CI for the median survival time using the method in K & M Chp. 4 (p. 120-122)

p <- 0.5
surv_tab %<>% mutate(linear = (surv-(1-p))/std.err,
                     loglog = ((log(-log(surv))-log(-log(1-p)))*surv*log(surv))/std.err,
                     asin = (2*(asin(sqrt(surv))-asin(sqrt(1-p)))*sqrt(surv*(1-surv)))/std.err)

surv_tab

surv_tab %>% filter(between(loglog,-1.96,1.96)) %>% select(time) %>% range()
surv_tab %>% filter(between(linear,-1.96,1.96)) %>% select(time) %>% range()
surv_tab %>% filter(between(asin,-1.96,1.96)) %>% select(time) %>% range()

#compare this to the 95% CI for the median that is produced when printing a summary of the survfit object (or using quantile.survfit). Are the intervals the same?

tongue_km
quantile(tongue_km, probs=0.5)

plot(tongue_km)
abline(h=0.5,col="red")
abline(v=c(42,104), lty=3, col="red") #from quantile
abline(v=c(42,100), lty=3, col="blue") #from log-log calc in book

```


```{r}
# Example 4.2 K&M p120-122
data(bmt)

p<-0.5
bmt_surv<-bmt %>% as.tibble() %>% filter(group==1) %$% Surv(t2, d3)
bmt_km<-survfit(bmt_surv~1, type="kaplan-meier",error="greenwood")
bmt_tab<-summary(bmt_km) %$% cbind(time, n.risk, n.event, n.censor, surv, std.err) %>% as.tibble()

bmt_tab %<>% mutate(linear=(surv-(1-p))/std.err,
                    loglog=((log(-log(surv))-log(-log(1-p)))*surv*log(surv))/std.err,
                    asin=(2*(asin(sqrt(surv))-asin(sqrt(1-p)))*sqrt(surv*(1-surv)))/std.err)
bmt_tab

bmt_tab %>% filter(between(linear,-1.96,1.96)) %>% select(time) %>% range()
bmt_tab %>% filter(between(loglog,-1.96,1.96)) %>% select(time) %>% range()
bmt_tab %>% filter(between(asin,-1.96,1.96)) %>% select(time) %>% range()
```

#### Estimators for left-truncated, right censored data

```{r}
# read in the psych data and create a survival object using 'ageentry' as the left truncation time
psych<-read_csv("psych.csv")

# get K-M estimates and log-log confidence intervals by sex 
psych_km<-survfit(Surv(ageentry, agestop, death)~sex, 
                  data=subset(psych,ageentry>=21), conf.type="log-log")
            
#plot conditional survival estimates given survival to age 21
autoplot(psych_km,xlim=c(21,80))
       
#plot conditional survival estimates given survival to age 35
psych_km35<-survfit(Surv(ageentry, agestop, death)~sex,
                    data=subset(psych,ageentry>=35),conf.type="log-log")
     
autoplot(psych_km35,xlim=c(21,80))

```
