---
title: BIOS 7323 Lab 5
date: 2017-10-01
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=file.path("/home/nathan/Dropbox/njames/school/PhD/courses/2018_19/ta_surv_anlys2/labs/lab5"))

#tidyverse pkgs
library(tidyverse)
library(GGally)
library(survminer)
library(magrittr)
library(scales)

#survival analysis pkgs
library(survival)
library(KMsurv)
```

### Objectives

Perform 1 and 2 sample hypothesis tests in R

#### One sample hypothesis tests

A review of the tongue cancer literature found two previous manuscripts that used parametric models for overall survival. The first paper used an exponential model with $\lambda=0.009$ (i.e. $S(t)=\exp(-0.009t)$) while the second paper used a log-logistic model with $\lambda=0.016$ and $\alpha=1$ ($S(t)=\frac{1}{1+0.016t}$).

We can test whether the data are consistent with a specific distribution by calculating the expected probability of survival at each time in our dataset and including these times as an offset term in the formula argument of survdiff().

```{r}
#load the tongue cancer data and create a plot that shows the K-M estimate for overall survival (ignoring tumor type) and the two parametric models from the previous papers:
# exponential with lambda = 0.009, 
# log-logistic with lambda = 0.016, alpha = 1
data(tongue)


# perform 2 one sample log-rank tests to see if the data is consistent with the previous parametric models.
# Hint: use offset(expected_survival) on the right side of the survdiff() formula

# test comparing to exponential with lambda = 0.009 


# test comparing to log-logistic with lambda = 0.016, alpha = 1


```

Are the observed data consistent with the parametric models used in the two papers?

#### 2+ Sample hypothesis tests with no covariates

In the bone marrow transplant (bmt) data, suppose investigators are interested in the association between receiving methotrexate (MTX) as a graft-vs-host prophylactic and time to either death or relapse (disease-free survival).

```{r}
# use the bmt data to create a dataset with event time and indicator for disease-free survival as well as MTX treatment status and disease group (ALL, AML low risk, AML high risk)
data(bmt) 


# plot Kaplan-Meier disease-free survival curves for each treatment group
# make sure to add a legend or label to identify the groups


```

Now test the hypothesis that the hazard rate is same for both MTX treatment groups at all time points vs. the alternative that the hazard rates differ at some time point.

```{r}
# Use survdiff() to perform a log-rank test of the hypothesis that the hazard rate is same for both treatment groups


# test the same hypothesis using the Peto-Peto modification of the Gehan-Wilcoxon test


```

What is the interpretation of the two test results? How do the tests differ in terms of the weights given to each time point?

#### 2+ Sample hypothesis tests adjusting for covariates

Before bone marrow transplantation, the investigators grouped patients into three risk categories: ALL, AML low risk, and AML high risk. Suppose we want to adjust for the patient risk category when testing for association between MTX treatment and disease-free survival.

Later, you will discuss how to adjust using regression modeling. Today, we will use stratification.

```{r}
#plot Kaplan-Meier disease-free survival curves for each of the 6 treatment/risk group categories


# Test whether the hazard rate between the MTX treatment groups differs when stratifying by the risk level


```

What conclusion do you draw from this stratified test?
