---
title: BIOS 7323 Lab 3
date: 2018-09-17
output:
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=file.path("/home/nathan/Dropbox/njames/school/PhD/courses/2018_19/ta_surv_anlys2/labs/lab3"))

#tidyverse pkgs
library(tidyverse)
library(ggfortify)
library(magrittr)

#survival analysis pkgs
library(survival)
library(KMsurv)
library(OIsurv)
```

### Objectives

Calculate and compare non-parametric estimators of Survival (S(t)) and Cumulative Hazard (H(t)) for right-censored data

Construct pointwise confidence intervals and confidence bands for the Survival function

####  Non-parametric survival estimators

##### Kaplan-Meier

```{r}
#load drug 6mp data
data(drug6mp)

# create a Surv object using time=t2 and event=relapse

# use survfit() to get a Kaplan-Meier (K-M) estimate of survival for 6mp patients
# hint: check documentation for survfit.formula 


# summarize the survfit object using summary()


# compare the structure of the survfit object and its summary


# find the median survival time


# plot the K-M curve


```

##### Aside: Survival estimates for multiple groups

```{r}
# load tongue cancer data
data(tongue)

#get K-M estimates and plot curves by tumor type using formula notation
# i.e. survfit(Surv(time,event) ~ type, ...)

```

##### Fleming-Harrington

```{r}
# use survfit to get a Fleming-Harrington (F-H) estimate of survival for 6mp patients


# find the median survival time based on the F-H estimate


# compare the K-M estimate to the F-H estimate by plotting both on the same graph


```

####  Non-parametric cumulative hazard estimators

```{r}
#use the relationship H(t) = -log(S(t)) to calculate the cumulative hazard using both the K-M and F-H survival estimates from above


# Use the calcna() function below to directly calculate the Nelson-Aalen estimate of cumulative hazard
# (http://sas-and-r.blogspot.com/2010/05/example-739-nelson-aalen-estimate-of.html)

calcna = function(time, event) {
   na.fit = survfit(coxph(Surv(time,event)~1), type="aalen")
   jumps = c(0, na.fit$time, max(time))
   # need to be careful at the beginning and end
   surv = c(1, na.fit$surv, na.fit$surv[length(na.fit$surv)])

   # apply appropriate transformation
   neglogsurv = -log(surv)   
   
   # create placeholder of correct length
   naest = numeric(length(time))  
   for (i in 2:length(jumps)) {
      naest[which(time>=jumps[i-1] & time<=jumps[i])] = 
         neglogsurv[i-1]   # snag the appropriate value
   }
   return(naest)
}


# Compare the cumulative hazard estimates calculated from the K-M and F-H survival estimates with the Nelson-Aalen estimate of cumulative hazard
# What can you say about the 3 estimates?


```

#### Pointwise confidence intervals for S(t)

```{r}
# The default confidence interval for survfit() is "log"
# get a new K-M estimate for the 6mp data using conf.type="log-log" instead
# Compare the "log-log" CI and the "log" CI for the K-M estimate

```


#### Confidence bands for S(t)

```{r}
# Use the tongue cancer data to create a survival object for time to death (ignoring type)
# and use this new Surv object to create a plot of the K-M survival estimate with:
# 1) pointwise "log-log" confidence intervals
# 2) equal probability (EP) confidence bands using log-log transform
# 3) Hall-Wellner confidence bands using log-log transform
# hint: use OIsurv::confBands() for 2) and 3)  

````
