---
title: BIOS 7323 Lab 3
date: 2018-09-17
output:
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=file.path("/home/nathan/Dropbox/njames/school/PhD/courses/2018_19/ta_surv_anlys2/labs/lab3"))

#tidyverse pkgs
library(tidyverse)
library(ggfortify) #to plot survival curves with ggplot2
library(magrittr) # to get %$% operator

#survival analysis pkgs
library(survival)
library(KMsurv)
library(OIsurv)
```

### Objectives

Calculate and compare non-parametric estimators of Survival (S(t)) and Cumulative Hazard (H(t)) for right-censored data

Construct pointwise confidence intervals and confidence bands for the Survival function

####  Non-parametric survival estimators

##### Kaplan-Meier

```{r}
#load drug 6mp data
data(drug6mp)

#create a Surv object using time=t2 and event=relapse
surv6mp <- as.tibble(drug6mp) %>% arrange(t2) %$% Surv(time=t2, event=relapse)
  
surv6mp_2 <- as.tibble(drug6mp) %>% arrange(t2) %>% {Surv(time=.[["t2"]], event=.[["relapse"]])}

identical(surv6mp, surv6mp_2)

# use survfit() to get a Kaplan-Meier (K-M) estimate of survival for 6mp patients
# hint: check documentation for survfit.formula 
sfit6mp_km <- survfit(surv6mp~1, type="kaplan-meier")

# summarize the survfit object using summary()
sfit6mp_km_summ <- summary(sfit6mp_km)

# compare the structure of the survfit object and its summary
str(sfit6mp_km)
str(sfit6mp_km_summ)

# find the median survival time
sfit6mp_km_summ$table["median"]
sfit6mp_km

# plot the K-M curve
plot(sfit6mp_km, mark.time = TRUE)

# using ggplot2 and ggfortify
autoplot(sfit6mp_km)

#Tatsuki Koyama's kmplot()
#source("http://biostat.mc.vanderbilt.edu/wiki/pub/Main/TatsukiRcode/RFunctions1.R")
#kmplot(sfit6mp_km)
```

##### Aside: Survival estimates for multiple groups

```{r}
# load tongue cancer data
data(tongue)

#get K-M estimates and plot curves by tumor type using formula notation
# i.e. survfit(Surv(time,event) ~ type, ...)
# base R
tongue_km <- survfit(Surv(time,delta)~type, type="kaplan-meier", data=tongue)
plot(tongue_km, lty=1:2, mark.time = TRUE)

# ggplot
survfit(Surv(time,delta)~type, type="kaplan-meier", data=tongue) %>% autoplot(conf.int=FALSE)
```

##### Fleming-Harrington

```{r}
# use survfit to get a Fleming-Harrington (F-H) estimate of survival for 6mp patients
sfit6mp_fh <- survfit(surv6mp~1, type="fleming-harrington", conf.int=FALSE)
summary(sfit6mp_fh)

# find the median survival time
sfit6mp_fh

# compare the K-M survival estimate to the F-H estimate by plotting both on the same graph
plot(sfit6mp_km, conf.int = FALSE)
lines(sfit6mp_fh, col="red")
legend("bottomleft", legend=c("Kaplan-Meier","Fleming-Harrington"), lty=c(1,1), col=c("black", "red"))
```

####  Non-parametric cumulative hazard estimators

```{r}
#use the relationship H(t) = -log(S(t)) to calculate the cumulative hazard using both the K-M and F-H survival estimates from above

hfit6mp_km <- -log(sfit6mp_km$surv)
hfit6mp_fh <- -log(sfit6mp_fh$surv)
cumhaz1 <- list(time=sfit6mp_km$time, chaz_km=hfit6mp_km, chaz_fh=hfit6mp_fh) %>% as.tibble()

# Use the calcna() function below to directly calculate the Nelson-Aalen estimate of cumulative hazard
# (http://sas-and-r.blogspot.com/2010/05/example-739-nelson-aalen-estimate-of.html)

calcna = function(time, event) {
   na.fit = survfit(coxph(Surv(time,event)~1), type="aalen")
   jumps = c(0, na.fit$time, max(time))
   # need to be careful at the beginning and end
   surv = c(1, na.fit$surv, na.fit$surv[length(na.fit$surv)])

   # apply appropriate transformation
   neglogsurv = -log(surv)   
   
   # create placeholder of correct length
   naest = numeric(length(time))  
   for (i in 2:length(jumps)) {
      naest[which(time>=jumps[i-1] & time<=jumps[i])] = 
         neglogsurv[i-1]   # snag the appropriate value
   }
   return(naest)
}

hfit6mp_na <- calcna(drug6mp$t2,drug6mp$relapse)


cumhaz2 <- list(time=drug6mp$t2,chaz_na=hfit6mp_na) %>% as.tibble() %>% distinct() %>% arrange(time)

# Compare the cumulative hazard estimates calculated from the K-M and F-H survival estimates with the Nelson-Aalen estimate of cumulative hazard
# What can you say about the 3 estimates?

left_join(cumhaz1, cumhaz2, by="time") %>% gather(type, cumhaz, -time) %>% ggplot(aes(x=time,y=cumhaz,color=type)) + geom_step() + facet_grid(.~type)
```

#### Pointwise confidence intervals for S(t)

```{r}
# The default confidence interval for survfit() is "log"
# get a new K-M estimate for the 6mp data using conf.type="log-log" instead
# Compare the "log-log" CI and the "log" CI for the K-M estimate

sfit6mp_km2 <- survfit(surv6mp~1, type="kaplan-meier", conf.type="log-log")

plot(sfit6mp_km, conf.int = TRUE)
par(new=TRUE)
plot(sfit6mp_km2,conf.int = TRUE,col="red")
```


#### Confidence bands for S(t)

```{r}
# Use the tongue cancer data to create a survival object for time to death (ignoring type)
# and use this new Surv object to create a plot of the K-M survival estimate with:
# 1) pointwise "log-log" confidence intervals
# 2) equal probability (EP) confidence bands using log-log transform
# 3) Hall-Wellner confidence bands using log-log transform
# hint: for 2) and 3) use OIsurv::confBands() 
survtong <- tongue %$% Surv(time, delta) 

tongue_km_all <- survfit(survtong~1, type="kaplan-meier", conf.type="log-log")
cb_ep <- OIsurv::confBands(survtong, confType = "log-log", type="ep")
cb_hall <- OIsurv::confBands(survtong, confType = "log-log", type="hall")

plot(tongue_km_all, mark.time = TRUE)
lines(cb_ep, col="green")
lines(cb_hall, col="red")
legend("topright",legend=c("pointwise (log-log) CI","EP CB","Hall-Wellner CB"),
       lty=c(2,1,1),
       col=c("black","green","red"))
````
